---
title: "New statistics for the design researcher"
author: "Martin Schmettow"
date: "April 3, 2017"
output:
  word_document:
    toc: yes
  html_document:
    fig_caption: yes
    number_sections: yes
    theme: journal
    toc: yes
bibliography: Bayesian_Stats.bib
---

<style type="text/css">
.table {
     width: auto;

}
</style>



```{r setup, 	message = FALSE,	warning = FALSE,	include = FALSE}
## The following is for running the script through knitr
purp.mcmc = F
source("RMDR.R")
# library(printr)1
CE <- c("IPump", "Overdisp")
load_CE(CE)

#formals(stan_glm)$iter <- 200
#formals(stan_glm)$chains <- 1

```


# Generalized Linear Models

Have you wondered about the abundance of simulated data sets up to this point? The reason for using simulated data is: the linear model, as introduced by now, makes assumptions that are *never* met by real data. The simulated data sets so far demonstrated some features found in real data sets, but  unsparingly wiped over some other frequent peculiarities. 

Up to this point, another question is probably lurking in the minds of readers with some classic statistics training:  what has happened to the assumptions of ANOVA and the like and where are all the neat tests covered that check for normality, constant variance and such?  In the next section we will review these assumptions and lead them ad absurdum. Simply put, these assumptions cannot never be true for any data and checking assumptions on model that you know is inapproriate is a futile futile exercise. This perspective holds at least, when better alternatives are available, and that is the case: with *Generalized Linear Models* we again extend the regression modelling framework. 

The GzLM framework rests on two extensions that bring us a huge step closer to the data.  The first one is a minor mathematical tweak, the *link function*. The second is the informed choice from a set of statistical distributions and this will open a whole new world for you. Follow up on this path and discover the many sources of randomness.

In the preceding chapters we got acquainted with the linear model as an extremely flexible tool to represent dependencies between predictors and outcome variables. We saw how factors and covariates gracefully work together and how complex research designs can be captured by multiple crossed random effects. It was all about specifying appropriate (and often sophisticated) right-hand sides formula, the predictors term. Little space has been dedicated to the outcome variables. That is now going to change, and we will start by examining the assumptions that are associated with the outcome variable.

In the following three sections I will explain the three core assumptions of linear models, recall the canonical model formulation:

$$μ_i=β_0+ β_1 x_1i+ …+β_k x_ki\\
y_i \sim Norm(μ_i,σ)$$

The first term, we called the likelihood. It represents the quantitative relations we expect to find in the data. When it is a sum of products, like above, we call it linear. *Linearity* is the first assumption that fails. The second term represents the source of randomness and it hosts two further assumptions: *normal distribution* of errors and *constant error variance*. 

Some classic textbooks tend to present these assumptions as preconditions for a successful ANOVA or linear regression. As in pre-condition, they need to be checked upfront using a zoo of tests. If one fails, let's say the Kolmogorov-Smirnoff test on normality, one better considers non-parametric tests or a down-tempering statement. Here, I will not present any tests on assumption, other than exploratory plots. Instead, I will right-away debunk all three assumptions for measures that design researchers are routinely dealing with. The three debunks take the LM formula  literally by simulating data from it. The *simulations* will raise severe skepticism as they do not look like real data. An idea as equally simple and immensely useful as the liner model can expected to be a tough survivor. Less useful ideas have outlived their expectation for decades and centuries. Therefore, in this chapter, *real data* will speak for itself, the infusion pump case study (*Ipump*).  After all, the poor fidelity of linear models, aka the  *Gaussian GzLM* is the real reason why the previous chapters made regular use of simulated, but not real, data.

Every assumption that fails, rebounces from one of two building blocks that GzLM add to our regression framework:

1. *link functions* re-establish linearity
2. *random distributions* cover randomness of "real" outcome variables

After a full debunk, the *GzLM* framework will rise from the ashes of *LM*. It hosts a variety of models that leaves little reason for crude approximations of the linear model, aka the *Gaussian LM*. The *Poisson LM* is the logical choice for outcome variables that are counted with no limit, like number of errors.  *Binomial regression* covers the case of succesful task completion, where counts have an upper boundary. These two GzLM members have been around for more than half a century in statistics. The quest for a good model for time-on-task was more difficult as the seemingly good candidates disappointed at first. Luckily, the with recent developments in Bayesian regression engines the choice of random distributions has become much broader. For ToT, I here suggest the exponentially-modified Gaussian *(ExGauss) LM*. Also for analogue rating scales will I present a relative novelty, the *Beta LM*.

## Case: user testing infusion pumps

Medical infusion pumps are unsuspicious looking devices that  are en-mass installed in surgery and intensive care. Their only purpose is controlled injection of medication in the blood stream. Pumps are rather simple devices as infusion is a function of volumen and time, only. They are routinely used by trained stuff, anaesthesiologists and  nurses. We should have great faith in safe operation under such conditions. The truth is, medical infusion pumps have reportedly killed dozens of people, thousands were harmed and an unknown number of nurses lost their jobs. The past generation of pumps is cursed with a hilarious set of completely unnecessary design mistakes:

* tiny 3-row LCD displays
* flimpsy foil buttons without haptic marking or feedback
* modes
* information hidden in menus

For fixing these issues no additional research is needed, as the problems and there fixes are pretty obvious to experienced usability specialists. What needs to be done, though, is proper validation testing of existing and novel interfaces, for example:

* is the interface safe to use?
* is it efficient to learn?
* is a novel interface better than a legacy design? And by how much?

We conducted such a study. 

[DESCRIBE STUDY]

## Debunking the Gaussian linear model

### Assuming linearity

Recall the principle, "all-finite-in-the-endless": when two or more interventions improve the same process, e.g. visual recognition of letters, the sum is less than the summands. This results in a non-linearity when the boundary of performance is reached. With a small set of predictors this can gracefully be modelled as saturation interaction effects.

Consider a study that assesses the improvement of safe operation with continued practice. For simplicity, we regard just a single nurse whose number of errors were measured on a chain of 8 tasks. Errors in operation were measured as number of deviations from the shortest possible interaction sequence. 

We simulate a linear model, assuming there is an improvement of one error less with every repetition of the sequence, with expected 7 deviations in the first session. We assume that path deviations has a normally distributed random component with $\sigma = 1$. 

```{r}
data_frame(session = as.integer(1:5),
           mu = 9 - session * 2.5,
           deviations = rnorm(5, mu, sd = 1)) %>% 
  ggplot(aes(x = session, 
             y = deviations)) +
  geom_point() +
  geom_abline(aes(intercept = 9, slope = -2.5))


```

See what happens when the linear model is naivly applied to the deviation counts. In no time, negative values are drawn, which is impossible. Similar to saturation effects, we would expect some asymptotic behaviour, rather than a straight line, when the outcome variable approaches its natural lower boundary. 

Here is some real world data from the IPump study. We take a look at a small slice of it: the total number of deviations of one participant across the three sessions.



```{r}

attach(IPump)

D_pumps %>% 
  filter(Part == 5, Design == "Novel") %>% 
  group_by(Part, session) %>% 
  summarize(deviations = sum(deviations)) %>% 
  ggplot(aes(x = session, y = deviations)) +
  geom_point() +
  geom_line() + 
  ylim(0,7)

detach(IPump)
```

What really happens, when a performance measure approaches its natural limit, is an asymptotic leaning-on. Neither will the line break through the limit, nor will it stop there abruptly.

Such a non-linearity happens to all outcome variables that have natural lower or upper boundaries, and that includes all outcome variables in the universe, except its very own spatial extension. All outcome variables in design research suffer from the problem of impossible predictions as a consequence of their boundedness:

* Errors and other countable incidences are bound at zero
* ToT is bounded at zero and probably also bound above when users loose their patience
* Rating scales are bound at the lower and upper extreme item
* Task completion has a lower bound of zero and upper bound is the number of tasks. Or the average task completion is in the range $[0,1]$






### Assumping normal distribution


The second term of a linear model, $y_i \sim Norm(\mu_i, \sigma)$ states that the observed values are drawn from normal distributions. Note that the plural  (distributions) is by intention. Two observed values $y_i$ and $y_j$ are only drawn from the same distribution, when they have the same expected value $\mu_i = \mu_j$. Normal distribution sounds like this were the regular case. In fact, the normal distribution is a reasonable approximation in a limited number of situations. 

Consider a study comparing a novel and a legacy interface design for medical infusion pumps. The researchers let trained nurses perform a single task on both devices and count the errors. Assuming, the average number of errors per tasks is $\mu_L = 3$ for the legacy device and $\mu_N = 1.2$ for the novel device, with standard deviation of $\sigma = .8$. We can simulate a basic data set as:

```{r opts.label = "rtut"}
N = 80
pumps_2 <- 
  data_frame(Design = rep(c("L", "N"), N/2),
             mu = if_else(Design == "L", 3, 1.2),
             errors = rnorm(N, mu, sd = 1))
```

We illustrate the data set using histograms:

```{r opts.label = "rtut"}
pumps_2 %>% 
  ggplot(aes(x = errors)) +
  facet_grid(~Design) +
  geom_histogram(bins = 20) +
  geom_vline(col = "red", xintercept = 0) +
  coord_flip()
```

When instantiated by simulation, the linear model looses its unsuspicious flare (just like medical infusion pumps themselves). A substantial number of simulated observations is *negative*, which strictly makes no sense for error counts. The pragmatic and impatient reader may suggest to adjust the standard deviation (or move the averages up) to make negative values less unlikely. That would be a poor solution for the following two reasons. First, the normal distribution supports the full range of real numbers. There is always a chance of negative simulations, as tiny as it may be. Repeatedly hitting keys until `pumps` contains exclusively positive numbers (and zero), obviously is poor practice. The second reason is that the simulations very purpose was to express and explore expectations from the linear model (CG). We can simply conclude that any model that assumes normally distributed errors must be wrong when the outcome is bounded below or above, which means: always.

Recall how linearity is gradually bended when a magnitude approaches its natural limit. A similar effect occurs for distributions. Distributions that respect a lower or upper limit tend to get squeezed like chewing gum into a corner. And that is true for the distribution of real data as well as the theoretical distributions that are truly edaquate to present the data. Only the normal distribution stoically retains its symmetry in all infinity. The figure below shows the distribution of 



### Assuming constant variance





## Concepts of Generalized Linear Models

GzLM is a *framework for modeling* that produces a *family of models*. Every member of this family uses a specific link functions and chooses a particular random distribution. Sometimes GzLM are confused as a way to relax assumptions of linear models, (or even called non-parametric).  They absolutely are not! Every member of its own makes precise assumptions  on the level of measurement and the shape of randomness (see Table A). One can even argue that Poisson, Binomial and exponential regression are stricter than Gaussian, as they use only one parameter, with the consequence of an even  tighter association between variance and expected value. A few members of GzLM are classic: Poisson, Binomial (aks logistic) and exponential regression have routinely been used before they were united  under the hood of GzLM. These and a few others are called *canonical* GzLM, as they possess some convenient mathematical properties, that made efficient estimation possible, back in the days of expensive computer time. 


### Re-linking linearity

The strength of the linear term (the likelihood) is its endless versatility in specifying relations between predictor variables and outcome. Unfortunately, it represents all associations as straight lines. These lines extend from $-\infty$ to $\infty$ and will cross the lower or upper boundaries of every known outcome variable. The model could predict what cannot happen.

Generalized linear models use a simple mathematical trick to keep the linear term, but confine the expected values. We cannot just cut it off,  But it can be bended. In arithmetics an abundance of functions exists for every possible purpose. The  *exponential* function supports the whole range of real numbers and bends it up into the non-negative range. Its inverse, the *logarithm* does the opposite: It restretches strictly positive numbers to the real range.

Other variables, like success rates or analog rating scales, have two boundaries. A function that bends the real range into $[0;1]$ is the *logistic* function. The two-sided bending makes its characteristic sigmoid shape. Again there is an inverse function, the log_odds, or *logit*. 


It could translate a strictly positive $\mu$

```{r, opts.label = "fig.wide"}
data_frame(x = seq(-3, 3, .01)) %>% 
  ggplot(aes(x)) +
  stat_function(fun = exp, aes(color = "exponential"))

data_frame(x = seq(-3, 3, .01)) %>% 
  ggplot(aes(x)) +
  stat_function(fun = inv_logit, aes(color = "logistic"))
```


In GzLM, a layer is drawn between expected value $\mu$ and the linear term, *linear predictor $\theta$*. The *link function* transforms between $\mu$ and $\theta$. For example, when $\y_i$ is a count variable, such that $\mu$ is bounded at zero, the logarithm makes the transformation to th unbounded $\theta$. The transformation bends the scale, but is monotonous, leaving the order intact. The back-transformation to the observed scale is called the *mean function*  













Using the link function comes at a cost: the linear coefficients $\beta_i$ no longer  have a natural interpretation, like "movin one step on the predictor, let the outcome change by $\beta_i$". The linear link is now strictly with the linear predictor $\theta_i$. Still, we will see that other, only slightly less intuitive statements can 
be made for link-transformed linear models.

Who needs a well-defined link between observations and expected values? Applied design researchers do when predictions over time is their business. Reconsider Andrew & Jane in [WIS], who predicted the population mean of ToT from a sample. This is a statement on the status quo, and therefore it is not a prediction, in a strict sense. Comparing designs A and B is status quo, too, as both designs do exist. Generally, we  classify all research questions as *status quo* that stay confined in the range of $y_i$. In the IPump study it is compelling to ask: "how will the nurse perform in session 4?". "When will she reach error-free operation?". These are predictive questions, where the question reaches into the unknown.   

With Bayesian regression engines we can compute a best guess for observed ranges and factor levels, but we can also simulate in the void. We estimate a linear model on learning 

```{r stanglm_predict}
attach(IPump)
# M_1 <- 
  

```


```{r predicting_the_impossibe}
detach(IPump)
```




Some researchers apply transformations to their variables, to make the meet the assumptions. 


### Fun with random distributions 

Count measures are bounded to the below. Unlike the normal distribution, the minimal count is 0. Theoretically, there is no upper bound for counts, but larger counts just get more and more unlikely. The Poisson distribution describes count data. Different to the successes-on-trials, these counts have no natural upper limit. The distribution is left-skewed. The smaller the measured values are, the stronger the skew. Vice versa, when the average count gets larger, the distribution becomes more symmetric. A common misconception is that random distributions approach the normal distribution with larger sample sizes. That is not true. Increasing the number of participants just results in a cleaner shape of the distribution, with less ruggedness. Examine yourself, how the shape of Poisson distributions changes by mean count, as well as sample size. The following code simulates a Poisson distribution with a given $\lambda$ and $N$. With every new run of the code, `rpois` generates a new random set of counts. By repeatedly running this code and watching the output, you can get a good idea of how varied the shape of the distribution can be.

```{r fig:exercise_poisson_dist}
data_frame(count = rpois(n = 20,
                         lambda = 2)) %>% 
  ggplot(aes(x = count)) + 
  geom_histogram()

```



```{r fig:poisson_dist, opts.label = "future"}
expand.grid(lambda = c(2, 5, 10), N = 10, 50, 100) %>% 
  as_data_frame() %>% 
  mutate()
ddply()
```






## Count data


### Binomial regression

Now we simulate a data set with the same research design, but the outcome variable is successes in ten tasks. For instance, the designs are three websites and participants are given ten information retrieval tasks.

Because the research design is the same, we can keep the two object data frames. The only issue is that $\beta$ and $\theta$ are on the scale of the linear predictor $\eta$, which is linked to expected values $\mu$ by the *logit link function*:

$$\eta = \log(\mu/(1-\mu))$$ 

Expected values $\mu$ are computed from $\eta$ through the so-called *mean function*, which is the inverse of the logit, the *logistic tranformation*:

$$\mu = \exp(\eta)/{1 + \exp(eta)}$$ 

Note that values of \mu from the Gaussian data set above, are in the range of 300 - 350, which would result in chances of success of virtually 100%. To get more realistic values, we first divide $\eta$ by 100. 

SNIPPET 

from the experiment on Visual EDA:

Logistic regression coefficients are on a logit scale. For the interpretation, we want to back-transform to the original scale, which is: probability the assumption to be rejected. In order to make statements on the probability scale, we backtransform using the inverse function of the logit, the logistic function:
mu_x = exp(eta_x)/(1 + exp(eta_x))
eta in the above formula is called the "linear predictor". The linear predictor is the value you get when you do operations on the coefficients. The most simple application is computing the overall probability for a rejection. This is represented by the Intercept parameter. In the table below, this parameter accounts to practically zero, on the logit scale. Back-transforming to the probability scale gives exactly a chance of 50%:
exp(0)/(1 + exp(0)) = 1/2
 So, an intercept of zero always means that chances are equal, which is the same as saying: the coefficient does not add any information. The same is true for parameters: the closer a coefficient is to zero, the less does it influence prediction of reject. Like with general linear models, parameters close to zero have "no effect" and can even be removed to simplify a model (we don't do that here).
Note, that the intercept in our model represents the situation that skew is 0 and N = 0. A more useful baseline rate for rejection rate would be at skew = 0 and the smallest sample size N = 10.
We can compute any prediction we want by combining the parameters with the predictors. But, the other coefficients in the model represent differences, not absolute values. Therefore, when predicting a certain event, let's say rejection at a skew = 0 and N = 50, we first have to compute the linear predictor eta, then transform to get predicted value mu. 
eta = 0.042 + Skew * 0 + Sampel size * 10
mu = logist(eta)
To give another example, the chance of rejection at skew = .5 and N = 50 is computed as:
eta = 0.042 + Skew * 0.5 + Sampel size * 50
mu = logist(eta)
Watch out! It is a severe mistake to transform coefficients individually, then do the linear combination.
eta = logist(Skew * 0.5) + logist(Sampel size * 50)
In any case, it is much preferred to report the estimates on the probability scale, as this is right-away interpretable. Credibility limits can be reported on the linear predictor scale. 


Group effects indicate how homogeneous the overall pattern (Table 1) is on individual units in the data set. Relevant units here are the participant and stimuli. For exanple, the relativels large coefficient Skew means that people differ a lot in how much their response is influenced by skew (which is desireable): The strong effect on stimuli means that (after taking skew and N into account) stimuli differ a lot in which response they provoke. It sounds a bit paradox, but that means that stimuli differ systematically: there exist properties other then Skew and N, that trigger a response.

SNIPPET

```{r binomial, opts.label = "future"}
n.subj = 50
D4 = expand.grid(Participant = as.factor(1:n.subj),
								 Design = c("A", "B", "C")) %>%
	join(D3_P, by = "Participant") %>% 
	join(D3_D, by = "Design") %>%
	mutate(eta = (theta + beta)/100) %>%   ## linear predictor eta
	mutate(mu = plogis(eta)) %>%  ## expected value mu
	mutate(trials = 10) %>% 
	mutate(Successes10 = rbinom(3 * n.subj, trials, mu)) ## observed variable

## Distribution of linear predictor eta
D4 %>% 
	ggplot(aes(x = eta, fill = Design)) +
	geom_density(alpha = .5) +
	facet_grid(Design~.)

## Distribution of expected value mu (or p)
D4 %>% 
	ggplot(aes(x = mu, fill = Design)) +
	geom_density(alpha = .5) +
	facet_grid(Design~.) +
	xlim(0,1)

## Distribution of (simulated) observed values
D4 %>% 
	ggplot(aes(x = Successes10, fill = Design)) +
	geom_histogram(alpha = .5) +
	facet_grid(Design~.)

M104 = MCMCglmm(cbind(Successes10,trials) ~ Design,
					family = "multinomial2",
					random = ~ Participant,
					data = D4)

summary(M104)

data.frame(summary(M104)$solution) %>% 
	add_rownames() %>% 
	mutate(mu_hat = plogis(post.mean))
```




### Poisson regression

When the outcome variable is a count with no upper limit, Poisson regression applies. With counts we usually expect the variance of randomness to raise with the mean. The Poisson distribution  is very strict in the sense that the variance equals the mean $$y_i \sim \textrm{Pois}(\lambda) \rightarrow \textrm{Var}(x_i) = \textrm{Mean}(x_i) = \lambda$$. Real data frequently has variance that raises proportionally with the mean, but is inflated. This is called overdispersion and is accounted for by an observation-level random effect, which will be explained in a separate section.

The link function is the logarithm, as it transfoms from the non-negative range of numbers to real numbers. The use of a link function (other than identity) has as a consequence that the parameters of Poisson regression cannot be interpreted as differences in units. As we will see now, there is a rather intuitive interpretation of linear coefficients in Poisson regression, that follows from:

$\exp(a + b) = \exp(a)\exp(b)$

Hence, the likelihood (including the mean function) of a Poisson model with two parameters can be re-written as a purely multiplicative term.

$$
y_i = \exp(\beta_0 + x_1\beta_1 + x_2\beta_2)\\
y_i = \exp(\beta_0) \exp(x_1\beta_1) \exp(x_2\beta_2)
$$
With the log link function, the exponentiated coefficients simply become multiplicators and statements of ratio are meaningful, like the following: *With the legacy design users make 1.5 times the errors than with the novel design*. This corresponds nicely with the measurement-theoretic perspective that count variables have a natural origin (zero) and therefore are invariant to rescaling.

In the infusion pump study, the deviations from normative path were counted to serve as a measure for safety of operation. For demonstration of how to report Poisson regression, the reduction of deviations by repeated sessions is examined. As we are interested in the improvement from first to second session and second to third, successive difference contrasts apply.

The table below shows the regression estimates on the linear scale and the exponentiated parameters. The latter are interpretable on the original scale, but strictly as multiplicators.

```{r}
attach(IPump)
```

```{r fit:deviations, opts.label = "mcmc"}
## Setting successive difference contrasts (library(MASS))
M_dev <- D_pumps %>%  
  stan_glmer(deviations ~ Design + session + session:Design + 
               (1 + Design + session|Part) +
               (1 + Design|Task) +
               (1|Obs), ## accounting for overdispersion
             data = ., family="poisson") 

```

```{r opts.label = "mcmc"}
## checking the factorial 

M_dev_fct <- D_pumps %>%  
  stan_glmer(deviations ~ Design + Session + Session:Design + 
               (1 + Design + Session|Part) +
               (1 + Design|Task) +
               (1|Obs), ## accounting for overdispersion
             data = ., family="poisson") 
T_dev_fct <- fixef(M_dev_fct)
```


```{r}
T_dev <- fixef(M_dev)
# %>% 
#   mutate(`exp(center)` = exp(center),
#          `exp(lower)` = exp(lower),
#          `exp(upper)` = exp(upper)) %>% 
#   discard_redundant() %>%
#   coef()
  

T_dev
```

```{r}
detach(IPump)

```


The legacy design serves as the reference group. Note that  the session variable has been coded starting at 0, such that the intercept coefficient represents the first session.  The expected number of deviations is $\exp(`r frm_coef(IPump$T_dev, row = 1, interval = F)`) = frm_coef(IPump$T_dev, row = 1, interval = F, mean_fnc = exp)$. 

[CONTINUE RESULTS]

The  transformation of coefficients to the original scale has been applied to the point and range estimates as produced by the `fixef` command. The less convenient alternative would have been to transform the values of the MCMC chain. This shortcut is not generally allowed. The crucial issue is that link and mean functions are monotonically increasing, with the consequence that the order of observations is preserved. Formally, for any two MCMC iterations $i$ and  $j$ for a parameter $\beta_i$:

$$\beta_{1i} \lt \beta_{1j} \rightarrow \exp(\beta_{1i}) \lt \exp(\beta_{1j})$$
Recall that throughout this book, center and interval estimates have been obtained by simple quantiles, marking the points where 2.5%, 50% and 97.5% of all iterations are smaller. As order does not change with monotonous transformations, quantiles do not either. Some researchers prefer the mode of the posterior to represent its center location. The mode is the point of highest density. As it does not rely on ranks, it is even invariant under all transformations that preserve identity. Without question, the mode is also not affected by the mean function and one may dor the transformation after summary.

This is different for higher order methods for obtaining point and interval estimates. Most notably the mean and the highest posterior density intervals are not generally invariant to mean functions. When using those, the mean function must be applied before summarizing the posterior, which is rather inconvenient.


### Overdispersion

Poisson and binomial distributions differ from the well-known Gaussian in that they only have one parameter. Location and variance are two completely independent choices for Gaussian distributions, whereas for variables with Binomial or Poisson distribution, only the location parameter is estimated. The subtle consequence is that the variance of any such distributions is strictly tied to the mean. In fact, for Poisson distributions the variance *is* the mean:

$x \sim Pois(\lambda) \implies \mu = \sigma^2 = \lambda$

The variance of binomial variables is:

$x \sim Bin(p, k) \implies \mu = k\mu(1-\mu) = p$, where k is the number of trials.

[ILLUSTRATION]

The strict variance assumptions of Poisson, binomial and exponential models are frequently violated by real data. The violation happens when real impact factors have not been included in the likelhood equation. Whenever there is one or more imoact factors on the outcome at question that the researcher has not regarded, overdispersion inevitably happens. That means it always happens in studies other than purified physics experiments. 

Using Poisson regression for overdispersed  count data results in inflated certainty of estimates. When strictly inferential statistics is done, this must be accounted for. A classic way to regard overdispersion is to choose yet another random distríbution that adds a scale parameter for randomness variation. For Poisson regression that is the negative binomial distribution (never mind its name!), for binomial the beta-binomial and for exponential it is the gamma distribution. These are respected members of the broader GzLM family and two of them, negative binomial and gamma, are even canonical. Negative binomial and gamma distributions introduce a scale parameter for variance. They keep the basic relation between $\mu$ and variance intact, but 

With the broad implementation of random effects models in Bayesian regression engines, there is a generic procedure to capture the extra variance without extending the family any further. The trick is to introduce an *observation-level random effect*. Recall how we regard variation between members of a population as normally distributed deviations from the population mean, by the example of a Poisson grand mean model with a participant-level ($p$) random effect:

$$
\theta_{pi} = \beta_0 + x_p\beta_{0p} \\
\mu_{pi} = \exp(\theta_{pi})\\
\beta_{0p} \sim N(\mu_{p}, \sigma_p)\\
y_{p} \sim \textrm{Pois}(\mu_{ij})$$

The OLRE is normally distributed but  does not cause any bounded-range range, as it is added on the level of the linear predictor, before applying the exponential transformation. Observation-level random effects are completely analogous, except that every observation becomes its own group, in a Poisson grand mean model with added variation:

$$
\theta_{i} = \beta_0 + \beta_{i} \\
\mu_i = \exp(\theta_i)\\
y_{ij} \sim \textrm{Pois}(\mu_{ij})$$

See, how $\beta_i$ is a unique deviation per observation $i$, and how a variance parameter $\sigma$ appears in an otherwise purely Poisson model. Observation-level random effects are on the linear predictor level, and therefore additive. Compare this to the negative binomial distribution where variance is scaled up, which is multiplication. We find a resemblance with how sums on the linear predictor become multiplications on the expected values scale.

```{r sync_env_overdisp, opts.label = "invisible"}
#try(detach(Overdisp))
#syncenv::new_syncenv(SE = "Overdisp")
attach(Overdisp)
```

For demonstration of the concept, we simulate from an overdispersed Poisson grand mean model with participant-level random effects, and recover it via regression. 



The simulation first generates the PLRE and OLRE

```{r sim:overdisp_2, opts.label = "sim"}

simulate_1 <- function(
  beta_0 = 2,   # mu = 8
  sd_Obs = .3, 
  sd_Part   = .4,
  N_Part = 30,
  N_Rep  = 20, 
  N_Obs =  N_Part * N_Rep,
  seed = 42){
  set.seed(seed)
  Part <- data_frame(Part = 1:N_Part,
                     beta_0p = rnorm(N_Part, 0, sd_Part)) ## participant-level RE
  D <- data_frame(Obs  = 1:N_Obs,
                    Part = rep(1:N_Part, N_Rep),
                    beta_0i = rnorm(N_Obs, 0, sd_Obs),   ## observeration-level RE
                    beta_0 = beta_0) %>%
    left_join(Part) %>% 
    mutate(theta_i = beta_0 + beta_0p + beta_0i,
           mu_i    = exp(theta_i),               ## inverse link function
           y_i     = rpois(N_Obs, mu_i))
  D
}

D_1 <- simulate_1()

D_1 %>%
  ggplot(aes(x = y_i)) +
  geom_histogram()
```

```{r opts.label = "mcmc"}
M_1 <- 
  D_1 %>% 
  stan_glmer(y_i ~ 1 + (1|Part) + (1|Obs), data = .,
             family = poisson)

RE_1 <- ranef(M_1)
RE_1 %>% 
  group_by(re_factor) %>% 
  summarize(sd(center))
```

Random effect variation is accurately recovered from the simulated data. The following two plots show, the participant latent scores can be recovered and even the observation levels themselves. Every observation gets an accurate measure of how much it had been pushed by unrecorded sources of variation. Practically, we obtain residuals which can be used for model criticism. For example, extreme outliers can be identified and relations between random variation and predicted values (or groups)  are open to scrutiny.


```{r}
T_1_grpef <- grpef(M_1)
T_1_grpef

D_1 %>% 
  bind_cols(filter(RE_1, re_factor == "Obs")) %>% 
  ggplot(aes(x = beta_0i, y = center, ymin = lower, ymax = upper)) +
  geom_errorbar()


D_1 %>% 
  distinct(Part, beta_0p) %>% 
  bind_cols(filter(RE_1, re_factor == "Part")) %>% 
  ggplot(aes(x = beta_0p, y = center, ymin = lower, ymax = upper)) +
  geom_errorbar()

```


In the real world of design research, regular dispersion in Poisson, binomial or exponential data is completely unrealistic and overdispersion needs to be embraced. Within the framework of GzLM, the solution is to switch to a more flexible dedicated random distribution. From LMM we borrow a surprisingly simple and general solution, observation-level random effects. Why should we add some twisted random distribution to our toolbox when familiar random effects can be reused? It is amazing how neatly the model recovered the `r Overdisp$N_Obs` levels of the OLRE. For every observation, we get an estimate that is basically a *residual*. These are very useful for model criticism. In the next section, we will use OLRE to diagnose a more wicked source of overdispersion that cannot be covered by OLRE, zero inflation.

Frequently, I reminded the reader to interpret parameters quantitatively, by translating their magnitude to statements of practical relevance. For random effects variance this isn't always straight forward. usually more difficult. One possible way is to make comparative statements on the sources of variance *"the variance due to individual differences exceeds all other sources of variation taken together"*. OLREs are on the same scale as all other random effects in the model, which makes it a good default reference. A non-default comparison of sources of variance is the one of Dennis Egan, that people cause more variance than designs do. With GzLMM, the marriage of LMM and GzLM this claim is testable, as we will see in [CASE EGAN].

```{r, ops.label = "invisible"}
detach(Overdisp)
```


### Case: Egan's claim


### Zero inflation

Imagine a study that examines the frequency of visits to the social media website Fakebook. While other researchers already set out to find predictive factors, like extrovert personality and the like, we are here interested in the frequency of daily use.  

```{r sim:ZI, ops.label = "rtut"}
attach(Overdisp)

simulate_2 <- function(
  beta_0  = 2.5,
  N_Part  = 120,
  p_zero  = .2, # proportion of non-users
  sd_Part = 0.8, # individual differences (lp scale)
  seed    = 23 # parameters passed on to simulate_1
){
  set.seed(seed)
  data_frame(Part = 1:N_Part,
             theta = rnorm(N_Part, beta_0, sd_Part),
             mu = exp(theta),
             is_user = rbinom(N_Part, 1, 1 - p_zero), 
             visits = 0 + is_user *rpois(N_Part, mu))
}

D_2 <- simulate_2()
D_2 %>% 
  ggplot(aes(x = visits)) +
  geom_histogram()

```




```{r fit:ZI, opts.label = "mcmc"}
M_2 <-
  D_2 %>% brm(visits ~ 1 + (1|Part) , family = zero_inflated_poisson, data = .)
M_2

```

```{r}
detach(Overdisp)
```



## Modelling time

```{r}
attach(IPump)
D_pumps %>% 
  ggplot(aes(x = Session, col = Design, y = ToT)) +
  geom_boxplot()

D_pumps %>% 
  group_by(Design, Session) %>% 
  summarize(mean(ToT)) %>% 
  spread(Design, `mean(ToT)`)
detach(IPump)
```



### Time-to-failure models (and how they fail for ToT)

#### Weibull

```{r fit:tot_weib, opts.label = "mcmc"}
attach(IPump)
## Setting successive difference contrasts (library(MASS))
M_tot_wbl <- D_pumps %>%  
  brms::brm(ToT ~ Design + session + session:Design + 
               (1 + Design + session|Part) +
               (1 + Design|Task) +
               (1|Obs), ## accounting for overdispersion
            family="weibull", data = .) 


detach(IPump)

save_CE("IPump")

```

```{r}
IPump$M_tot_wbl
```



### ExGaussian regression

```{r fit:tot_exg, opts.label = "mcmc"}
attach(IPump)
## Setting successive difference contrasts (library(MASS))
M_tot_exg <- D_pumps %>%  
  brms::brm(ToT ~ Design + session + session:Design + 
              (1 + Design + session|Part) +
              (1 + Design|Task), ## accounting for overdispersion
            family="exgaussian", data = .) 
detach(IPump)

M_tot_exg
fixef(M_tot_exg)

save_CE("IPump")

```

```{r}
IPump$M_tot_exg
```


```{r opts.label = "future"}
attach(IPump)
D_pumps <-
  D_pumps %>% 
  mutate(M_tot_wbl = predict(M_tot_wbl)[,1],
         M_tot_exg = predict(M_tot_exg)[,1])

detach(IPump)



```

```{r opts.label = "future"}
attach(IPump)
P_pumps <- 
  bind_rows(
    posterior(M_tot_wbl),
    posterior(M_tot_exg)
  )
detach(IPump)



```




## Modelling rating scales

### Gaussian regression

The classic linear model we have mercilessly debunked at the beginning of the chapter, is a member of the GzLM framework, with some rather unusual properties. It assumes normally distributed randomness, which implies a constant variance. In contrast to all other members so far, there is no relation between mean and variance of randomness. The link function simply is *identity* and is consequential for the problem of linearity and bounded ranges. The general take is that, whenever there is a theoretically justified and reasonably pragmatic choice, one better leaves the Gaussian alone. Here it makes its re-entrance for a type of outcome measures that are so problematic that a pragmatic choice is indicated.


### Beta regression


```{r}
save_CE(CE)
```





