\documentclass{article}
\title{Glossary for New Statistics for the Design Researcher}
\author{Steffi Olbrich}

\begin{document}
\maketitle

\section{Chapter 4: Getting started with R}
\underline{\textbf{Glossary of words}}

\begin{description}

\item [Argument]
\item [Boolean algebra]
\item [Boolean operator]
\item [Character]
\item [CRAN comprehesive R archive network] a network of ftp and web servers around the world that store identical, up-to-date, versions of code and documentation for R. You will mostly download packages from here.
\item [Coefficient]
\item [Console]
\item [Data frame]
\item [Data transformation]
\item [Dichotomous variable]
\item [Environment] in R stores data in dataframes, variables, etc. 
\item [Factor]
\item [Function] a command you give, that performs a certain action and sometimes gives a return value.
\item [Higher level function]
\item [Index]
\item [Integer]
\item [Iteration]
\item [Level]
\item [Library] a collection of precompiled routines that a program can use. You will use mostly functions defined in certain libraries. See also: Package.
\item [Logical object]
\item [Logical operation]
\item [Statistical model]
\item [NA]
\item [Numeric]
\item [Object]
\item [Object class]
\item [Operator]
\item [Outlier]
\item [Package] In other programming languages it is called library, in R the library contains many packages.
\item [Plot] 
\item [Polish notation]
\item [R] is a programming language for mostly statistical purposes.
\item [R Script] 
\item [Rstudio] a programming environment for R, which offers many features and is userfriendly.
\item [Seed]
\item [Statistic]
\item [Statistical computing]
\item [Syntax] the text you type in R to define and command things.
\item [Variable]
\item [Vector]

\end{description}

\noindent \underline{\textbf{Glossary of packages}}

I think it would be nice to have a short description of each of the packages used or mentioned as well. 
\begin{description}

\item [bayr]
\item [devtools]
\item [dplyr]
\item [foreign] 
\item [ggplot] part of tidyverse, a package for creating pretty and useful plots in R.
\item [gridExtra]
\item [haven]
\item [knitr]
\item [mascutils]
\item [openxlsx]
\item [printr]
\item [readr]
\item [readxl]
\item [rstanarm]
\item [stringr]
\item [tidyr]
\item [tidyverse] a collection of libraries for simple and (more) intuitive data management in R.

\end{description}
\end{document}